A Practical Optional Type System for Clojure

### 2012-10-22 dissertation/future-work.tex
Example for infer

\section{Multimethods}
% Problem - typing isa?
% - probably impossible to type higher order usages of isa?.
% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
% Statically proving at least one relevant dispatch method is present.
% Warn potential method conflicts.
% Suppress warnings, similar to prefer-method syntax

@@

>>>>

\section{Multimethods}
\label{sec:mmfuture}

\begin{verbatim}
Multimethods in Clojure are an idiomatic and often used feature.
It would be essential for an optional static type system for
Clojure to support
There are at least two hurdles to supporting multimethods.
\end{verbatim}

% Problem - typing isa?
% - probably impossible to type higher order usages of isa?.
% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
% Statically proving at least one relevant dispatch method is present.
% Warn potential method conflicts.
% Suppress warnings, similar to prefer-method syntax

@@

A Clojure record is a composite of datatypes and maps; it is a datatype
with fields that can be treated like a map (records support map operations like \lstinline|get|,
\lstinline|assoc|, and \lstinline|dissoc|).
Records have some quirks that may prove difficult to

### 2012-10-22 dissertation/implementation.tex
Example for infer

This chapter discusses the implementation of the Typed Clojure prototype type 
system\footnote{We refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter.}
in some detail concentrating on significant challenges that were identified and overcome.
Many aspects of Typed Clojure's design follows the implementation of Typed Racket,
which is reflected in this chapter.

@@

Typed Clojure and Typed Racket differ significantly in how they perform type checking.
Typed Racket is implemented as a language on the
Racket platform, which provides highly sophisticated and extensible macro facilities.
Interestingly, this allows Typed Racket to be entirely implemented with macros.
Instead, Typed Clojure is implemented as a library that utilises abstract syntax trees (AST)
generated by analyze\footnote{https://github.com/frenchy64/analyze}, a library I developed for this project.
This strategy follows common practice for Clojure projects, which favours providing modular libraries 
over modifying the language.

@@

\begin{itemize}
  \item type alias definitions
  \item var, protocol, datatype, and Java Class annotations
  \item Java method annotations, such as \lstinline|nilable-param| and \lstinline|non-nil-return|
\end{itemize}

@@

The interface to the bidirectional checking algorithm is \lstinline|typed.core/check|,
which takes an expression, represented as an AST generated from \emph{analyze}, and an optional expected type for
the given expression. If the expected type is present, the bidirectional algorithm \emph{checks}
that the expected type matches the actual type of the expression.
If the expected type is omitted, a type is instead \emph{synthesised} for the expression.
The algorithm is based on Pierce and Turner's  Local Type Inference~\cite{PT00}
and the implementation is similar in shape to Typed Racket's~\cite{TF08}, in that
one function with an extra ``expected type'' argument is preferred over two complementary
functions, one for checking and one for synthesis.

The main difference between Typed Clojure's and Typed Racket's bidirectional checking
algorithm is the representation for expressions. Typed Racket relies on pre-existing
Racket features like syntax objects for expression representation. Clojure instead
leans towards abstract syntax tree representation, despite its Lisp heritages.
In terms of the bidirectional check, the difference is mostly cosmetic.

@@

\begin{itemize}
  \item TopFilter and BotFilter represent the trivially true and trivially false propositions respectively.
  \item TypeFilter and NotTypeFilter represent a positive or negative association of a binding name
        to a type. The syntax \lstinline|(is t name)|
  \item AndFilter and OrFilter represent logical conjunction and disjunction of propositions.
\end{itemize}

Propositions can optionally carry \emph{path} information
represented by a sequence of \emph{path elements}, which signify which part
of the binding's type to update. For example, Typed Racket uses \emph{car} and \emph{cdr} path elements
to track which component of a cons type to update.
Paths are particularly useful in Typed Clojure. There are path elements for traversing heterogeneous map types (KeyPE),
inferring length information (CountPE), and \lstinline|first| and \lstinline|rest| paths for sequences.
These additions do not appear to introduce any significant new complexities related to paths.

@@

The polymorphic type inference algorithm is directly ported from Typed Racket with slight extensions
for bounded variables,
and is reminiscent to Pierce and Turner's Local Type Inference~\cite{PT00}.
A common entry point for inferring type variables to polymorphic function invocations
\lstinline|typed.core/infer|.

@@

\begin{itemize}
  \item \lstinline|X| is a map from type variable names
        to their bounds (representing the type variables in scope),
  \item \lstinline|Y| is a map from type variable names
        to their bounds (representing the \emph{dotted} type variables in scope),
  \item \lstinline|S| and \lstinline|T| are sequences of types of equal length
  \item \lstinline|R| is a type,
  \item \lstinline|expected| is the expected type for R, or the value \lstinline|nil|,
\end{itemize}

@@

@@

@@

Instead, we generate new propositions using a technique 
suggested by Tobin-Hochstadt~\cite{Tob12},
that follows from the occurrence typing calculus defined 
by Tobin-Hochstadt and Felleisen~\cite{TF10}.
First the filter set for functions are inferred as usual.
Then the ``then-proposition'' is applied to the environment and new propositions
are added to the filter set for each variable in the type environment that changes.
The same procedure is followed for the ``else-proposition''.

@@

\section{Non-hygienic Macro Expansion and Filters}
% need to remove bindings as they fall out of scope
% Need hygiene to ensure correct propagation of filters
%(fn [a]
%  (if (= a 1)
%    (let [a 'foo] ; here this shadows the argument, impossible to recover filters
%      a)          ; in fact any new filters about a will be incorrectly assumed to be the argument
%      false))

>>>>

This chapter discusses the implementation of the Typed Clojure prototype type 
system (we refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter)
in some detail concentrating on significant challenges that were identified and overcome.
Many aspects of Typed Clojure's design follow the implementation of Typed Racket,
which is reflected in this chapter.

@@

Typed Clojure and Typed Racket differ significantly in how type checking
is integrated into their programming environments.
Typed Racket is implemented as a language on the
Racket platform, which provides highly sophisticated and extensible macro facilities.
Interestingly, this allows Typed Racket to be entirely implemented with macros.
Instead, Typed Clojure is implemented as a library that utilises abstract syntax trees (AST)
generated by analyze~\cite{Analyze2012}, a library I developed for this project.
This strategy follows common practice for Clojure projects, which favours providing modular libraries 
over modifying the language.

@@

\begin{itemize}
  \item type alias definitions
  \item global variable, protocol, datatype, and Java Class annotations
  \item Java method annotations, such as \lstinline|nilable-param| and \lstinline|non-nil-return|
\end{itemize}

@@

The interface to the bidirectional checking algorithm is \lstinline|typed.core/check|,
which takes an expression, represented as an AST generated from \emph{analyze}, and an optional expected type for
the given expression. If the expected type is present, the bidirectional algorithm \emph{checks}
that the expected type matches the actual type of the expression.
If the expected type is omitted, a type is instead \emph{synthesised} for the expression.
The algorithm is based on Pierce and Turner's  Local Type Inference~\cite{PT00}
and the implementation is similar in form to Typed Racket's~\cite{TF08}, in that
one function with an extra ``expected type'' argument is preferred over two complementary
functions, one for checking and one for synthesis.

The main difference between Typed Clojure's and Typed Racket's bidirectional checking
algorithm is the representation for expressions. Typed Racket relies on pre-existing
Racket features like syntax objects for expression representation. Clojure instead
leans towards abstract syntax tree representation, despite its Lisp heritage.
In terms of the bidirectional checking, the difference is mostly cosmetic.

@@

\begin{itemize}
  \item TopFilter and BotFilter represent the trivially true and trivially false propositions respectively.
  \item TypeFilter and NotTypeFilter represent a positive or negative association of a binding name
        to a type. The syntax \lstinline|(is t name)| means a proposition that the binding called
        \lstinline|name| is of type \lstinline|t| (corresponding to TypeFilter).
        The syntax \lstinline|(! t name)| means a proposition that the binding called \lstinline|name|
        is \emph{not} of type \lstinline|t| (corresponding to NotTypeFilter).
  \item AndFilter and OrFilter represent logical conjunction of propositions 
    (written \lstinline[mathescape]|(& $\overrightarrow{p}$)|), and
    logical disjunction of propositions
    (written \lstinline[mathescape]{(| $\overrightarrow{p}$)}) for propositions \lstinline|p|.

\end{itemize}

Propositions can optionally carry \emph{path} information
represented by a sequence of \emph{path elements}, which signify which part
of the binding's type to update. For example, Typed Racket uses \emph{car} and \emph{cdr} path elements
to track which component of a cons type to update.
Paths are particularly useful in Typed Clojure. There are path elements for traversing heterogeneous map types (KeyPE),
inferring length information (CountPE), and \lstinline|first| and \lstinline|rest| paths for sequences.
These additions do not appear to introduce any major new complexities related to paths.

@@

The polymorphic type inference algorithm is directly ported from Typed Racket with slight extensions
for bounded variables,
and is directly based on Pierce and Turner's Local Type Inference~\cite{PT00}.
A common entry point for inferring type variables for polymorphic function invocations
is \lstinline|typed.core/infer|.

@@

\begin{itemize}
  \item \lstinline|X| is a map from type variable names
        to their bounds (representing the type variables in scope),
  \item \lstinline|Y| is a map from type variable names
        to their bounds (representing the \emph{dotted} type variables in scope),
  \item \lstinline|S| and \lstinline|T| are sequences of types of equal length,
        (usually the types of the actual arguments provided and the types of the parameters
        of the polymorphic function),
  \item \lstinline|R| is a result type (usually the return type of the polymorphic function),
  \item \lstinline|expected| is the expected type for R, or the value \lstinline|nil|,
\end{itemize}

@@

\lstinline|infer| is almost always used when invoking polymorphic functions.
\lstinline|clojure.core/constantly| is a polymorphic function with type \lstinline|(All [x y] [x -> [y * -> x]])|
(read as a function taking \lstinline|x| and returning a function that takes any number
of \lstinline|y|'s and returns \lstinline|x|).
For instance, \lstinline|((constantly true) 'any 'number)| results in the value \lstinline|true|.

Type checking the invocation \lstinline|(constantly true)|
calls \lstinline|infer| roughly like

\begin{lstlisting}
(infer {'x no-bounds 'y no-bounds}
       {}
       [(parse-type 'true)]
       [(make-F 'x)]
       (with-frees [(make-F 'x) (make-F 'y)]
         (parse-type '[y * -> x]))
       nil)
\end{lstlisting}

where the internal Typed Clojure bindings

\begin{itemize}
  \item \lstinline|no-bounds| is the type variable bounds with upper bound as \lstinline|Any|
    and lower bound as \lstinline|Nothing|,
  \item \lstinline|parse-type| is a function that takes type \emph{syntax} and converts it to a type (its argument must be quoted), 
  \item \lstinline|make-F| is a function that takes a name symbol and returns a type variable type of that name, and
  \item \lstinline|with-frees| is a macro that brings the type variables named in its first argument into scope in its second argument.
\end{itemize}

This returns a substitution that replaces occurrences of \lstinline|x| with \lstinline|true|
and \lstinline|y| with \lstinline|Any|.
This helps infer a result type of \lstinline|(constantly true)|
as \lstinline|[Any * -> true]|.

@@

\begin{verbatim}
  Say more about the changes and why so much work was involved. Or at least
  the remaining changes were more tedious than anything.
\end{verbatim}

@@

Instead, we generate new propositions using a technique 
suggested by Tobin-Hochstadt~\cite{Tob12},
that follows from the occurrence typing calculus defined 
by Tobin-Hochstadt and Felleisen~\cite{TF10}.
First the filter set for functions are inferred as usual.
To collect new propositions, the ``then-proposition'' is applied to the type environment (which maps
local bindings to types). Any types associated with bindings that are changed after this
are represented as new propositions, which are added to the ``then-proposition'' for this filter set.
The same procedure is followed for the ``else-proposition''.

@@

\section{Non-hygienic Macro Expansion and Filters}

% need to remove bindings as they fall out of scope
% Need hygiene to ensure correct propagation of filters
%(fn [a]
%  (if (= a 1)
%    (let [a 'foo] ; here this shadows the argument, impossible to recover filters
%      a)          ; in fact any new filters about a will be incorrectly assumed to be the argument
%      false))

### 2012-10-22 dissertation/intro.tex
Example for infer

Recently a number of languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers a simple form of optional
static typing that specifically do not affect runtime semantics.
Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
a well-known dynamic language.
Typed Racket~\cite{TF08,Tob10} goes further by offering safe interoperability
between typed and untyped modules by generating appropriate runtime assertions based
on expected static types.

@@

A function with uniform variable parameters can treat its variable parameter
as a homogeneous list. 
Strickland et al. attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
in a function type, where \emph{T} is some type, and we take an identical approach.
For example, \lstinline|+| in Clojure accepts any number of arguments
of type \lstinline|Number|, represented by the type \lstinline|[Number * -> Number]|.

>>>>

Recently a number of languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart~\cite{Dart2012} (created by Google) is dynamically typed but offers a simple form of optional
static typing that specifically do not affect runtime semantics.
Typescript~\cite{TypeS2012} (created by Microsoft) adds an optional type system to Javascript,
a well-known dynamic language.
Typed Racket~\cite{TF08,Tob10} (by Tobin-Hochstadt et al.) goes further by offering safe interoperability
between typed and untyped modules by generating appropriate runtime assertions based
on expected static types.

@@

A function with uniform variable parameters can treat its variable parameter
as a homogeneous list. 
Strickland et al.\ attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
in a function type, where \emph{T} is some type, and we take an identical approach.
For example, \lstinline|+| in Clojure accepts any number of arguments
of type \lstinline|Number|, represented by the type \lstinline|[Number * -> Number]|.

### 2012-10-22 dissertation/lit-review.tex
Example for infer

Strickland et al. invented a type system supporting variable-arity polymorphism~\cite{STF09}
a version of which is included in the current implementation of Typed Racket.
Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
which are used to check definitions and usages of functions with non-trivial rest parameters

@@

Pierce and Turner split \emph{local type argument synthesis} into two further
algorithms: bounded, and unbounded quantification~\cite{PT00}. 
Typed Racket 
supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Piece et al.
Scala supports bounded quantification with F-bounded polymorphism~\cite{CCHOM89},
basing its type argument synthesis on the bounded quantification algorithm.

@@

\emph{Colored} types contain extra contextual information, including the propagation direction
and missing parts of the type. They are generally useful
for describing ``information flow in polymorphic type systems with propagation-based type inference''~\cite{OZZ01}. 
If the scope of my research permits it, I plan to investigate
using colored types to solve the polymorphic higher-order-function limitation of Local Type Inference
(see listing \ref{lst:hts}).

@@

@@

\begin{verbatim}
Many advanced type systems provide support for ..
Forward reference to algo.monad experiment.
\end{verbatim}

@@

Many related components must come together in the design of a
static type system. Typed Racket achieves a satisfying balance of 
occurrence typing, local type inference and variable-arity polymorphism.
Similarly, Scala features F-bounded polymorphism, a class hierarchy
that is compatible with Java, and colored local type inference.

>>>>

Strickland et al.\ invented a type system supporting variable-arity polymorphism~\cite{STF09}
a version of which is included in the current implementation of Typed Racket.
Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
which are used to check definitions and usages of functions with non-trivial rest parameters

@@

Pierce and Turner split \emph{local type argument synthesis} into two further
algorithms: bounded, and unbounded quantification~\cite{PT00}. 
Typed Racket 
supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Piece et al.\ .
Scala supports bounded quantification with F-bounded polymorphism~\cite{CCHOM89},
basing its type argument synthesis on the bounded quantification algorithm.

@@

\emph{Colored} types contain extra contextual information, including the propagation direction
and missing parts of the type. They are generally useful
for describing ``information flow in polymorphic type systems with propagation-based type inference''~\cite{OZZ01}. 
Colored Local Type Inference is a candidate for

@@

After some investigation, typing multimethods with Typed Clojure is assigned as future work
(see section \label{sec:mmfuture}).

@@

Many advanced type systems provide support for \emph{higher kinds},
which are ``another level up'' from types.
For example, a type \lstinline|Number| is distinguished from a
\emph{type constructor} \lstinline|Monad| (see section \ref{sec:monads}), which is a type level \emph{function}
(similar to the difference between values and functions on the value level, but on the type level).

F-$\omega$ is a typed $\lambda$-calculus with support for higher kinds,
specifically type constructors (see Pierce~\cite{Pie02}).
Haskell~\cite{Mar10} distinguishes between types and type constructors,
the latter of which is an essential part of its monad library 
(see section \ref{sec:monads} for a monad library ported to Typed Clojure
that uses type constructors similarly).

@@

Many related components must come together in the design of a
static type system. Typed Racket achieves a satisfying balance of 
occurrence typing, local type inference and variable-arity polymorphism.
Scala features F-bounded polymorphism, a class hierarchy
that is compatible with Java, and colored local type inference.
Typed Clojure takes inspiration from these, and similar, projects.
